---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.6,
  fig.align = "center"
)
```

```{r libraries, message = FALSE, include = FALSE}
library(ggplot2)
library(ggtypst)
```

`ggtypst` brings Typst-powered text and math rendering to `ggplot2`. It is for
the moments when plain plot text is not expressive enough, but a full external
Typst or LaTeX workflow would be too heavy.

## Why ggtypst?

Base `ggplot2` text helpers are great for ordinary labels, but scientific and
technical plots often need more than plain strings. Common needs include rich
text styling, multi-line labels, inline math, display equations, and theme text
that looks better than `plotmath` output.

`ggtypst` solves that gap by compiling Typst source to SVG and inserting the
result back into `ggplot2`. That gives you three main benefits:

- Typst markup is expressive and compact
- Typst math looks high quality and works naturally beside text
- the output stays inside normal `ggplot2` layers and themes

Typst itself is a modern markup language and typesetting engine. In practice,
that means you write source text such as `*bold*`, `_italic_`, `#linebreak()`,
or `$sum_(i=1)^n x_i$`, and Typst turns it into polished rendered output.

## Installation

Install from GitHub for now:

```{r install, eval = FALSE}
install.packages("remotes")
remotes::install_github("Yousa-Mirage/ggtypst")
```

`ggtypst` builds a Rust backend during installation, so you need `rustc` and
`cargo` on your system. You do not need a separate local Typst or LaTeX
installation.

## First plot

A small but useful example is like this:

```{r first-plot}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = "grey50") +
  annotate_typst(
    typst_code = "*Hello from Typst!* #linebreak() $sum_(i=1)^n x_i$",
    x = 4.8,
    y = 32,
    size = 14,
    color = "#1D4E89"
  ) +
  labs(
    title = "A first ggtypst plot",
    x = "Weight (1000 lbs)",
    y = "Miles per gallon"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_typst(size = 16, face = "bold", colour = "#102A43"),
    axis.title.y = element_typst(size = 14, face = "italic", colour = "#1D4E89")
  )
```

This shows the core model of the package:

1. you write Typst source as a string
2. `ggtypst` renders it to SVG
3. the SVG is inserted back into the plot

## When to use annotate, geom, and element

Use the three helper families according to where the rendered content comes
from.

| Helper family | Best for | Typical use |
|---|---|---|
| `annotate_*()` | one manually placed label | a callout, note, or equation |
| `geom_*()` | one label per row of data | point labels, grouped labels, math annotations |
| `element_*()` | theme text | titles, subtitles, axis text, strips, legends |

If the content is tied to one fixed location, start with `annotate_*()`. If the
content belongs to your data, use `geom_*()`. If the content belongs to plot
metadata, use `element_*()`.

## R raw strings

Many Typst and LaTeX strings contain lots of quotes and backslashes. R raw
strings make those much easier to write.

Use the form `r"(... )"`:

```{r raw-strings, eval = FALSE}
r"(*bold text*)"
r"($sum_(i=1)^n x_i$)"
r"(\frac{a}{b})"
```

Why this matters:

- ordinary R strings often need escaping like `"\\frac{a}{b}"`
- raw strings let you write LaTeX-style math and Typst markup more directly
- they are especially handy for MiTeX examples

## Text, Typst math, and MiTeX math

`ggtypst` supports three related but different kinds of content.

### Plain Typst text

Use `annotate_typst()`, `geom_typst()`, or `element_typst()` when the label is
mostly text, even if it contains a little inline math.

### Native Typst math

Use `annotate_math_typst()`, `geom_math_typst()`, or `element_math_typst()` when
you want to write math directly in Typst syntax.

```{r typst-math-example, eval = FALSE}
annotate_math_typst(
  typst_math_code = r"(sum_(i=1)^n x_i)",
  x = 1,
  y = 1
)
```

### MiTeX-backed LaTeX math

Use `annotate_math_mitex()`, `geom_math_mitex()`, or `element_math_mitex()` when
your input is already in LaTeX math syntax. `ggtypst` converts it with MiTeX,
then evaluates the converted expression in an embedded Typst scope.

```{r mitex-example, eval = FALSE}
annotate_math_mitex(
  latex_math_code = r"(\frac{1}{2} + \sqrt{3})",
  x = 1,
  y = 1
)
```

## annotate_*(): one-off plot annotations

`annotate_*()` is the easiest place to start because it behaves like a normal
manual annotation layer.

```{r annotate-example}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = "grey65") +
  annotate_typst(
    typst_code = paste(
      "*Fuel economy* #linebreak()",
      "#text(fill: rgb(\"#1D4E89\"))[Typst text] +",
      "$integral_0^1 x^2 dif x = 1 / 3$"
    ),
    x = 4.9,
    y = 33,
    hjust = 1,
    size = 12,
    color = "#102A43"
  ) +
  annotate_math_mitex(
    latex_math_code = r"(\eta = \frac{mpg}{wt})",
    x = 4.9,
    y = 27.5,
    hjust = 1,
    size = 15,
    color = "#1D4E89",
    face = "bold"
  ) +
  theme_minimal(base_size = 12)
```

Use this family when:

- you already know the annotation position
- the label is not mapped from data
- you want to add one highlighted note or equation

## geom_*(): data-driven labels

`geom_*()` turns Typst strings into a true data layer. Each row gets its own
rendered label and can map aesthetics like `colour`, `angle`, `hjust`, and
`face`.

```{r geom-example}
labels <- data.frame(
  wt = c(1.84, 2.32, 5.25),
  mpg = c(33.9, 30.4, 15.0),
  label = c(
    r"(*Toyota Corolla* #linebreak() 33.9 mpg)",
    r"(*Fiat 128* #linebreak() nimble + efficient)",
    r"(*Maserati Bora* #linebreak() $v_8$ flagship)"
  ),
  colour = c("#2A6F97", "#3A86A8", "#7A1E48"),
  face = c("bold", "plain", "bold.italic")
)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = "grey80", size = 2) +
  geom_typst(
    data = labels,
    aes(wt, mpg, label = label, colour = colour, face = face),
    size = 11,
    nudge_y = 0.8,
    show.legend = FALSE
  ) +
  scale_colour_identity() +
  theme_minimal(base_size = 12)
```

Use this family when:

- labels come from rows in a data frame
- each label may have different styling
- you want Typst output to behave like other `ggplot2` geoms

## element_*(): Typst in themes

`element_*()` applies Typst rendering to theme text. This is how you get Typst
in plot titles, subtitles, axis titles, legend text, and facet strips.

```{r element-example}
ggplot(mtcars, aes(wt, mpg, colour = factor(cyl))) +
  geom_point() +
  labs(
    title = r"(`ggplot2` + Typst-rendered theme text)",
    subtitle = r"(Titles and legends can use *Typst markup* directly)",
    x = r"($sum_(i=1)^n c_i$)",
    y = r"(\eta = \frac{mpg}{wt})",
    colour = r"(*Cylinders*)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_typst(size = 16, face = "bold", colour = "#102A43"),
    plot.subtitle = element_typst(size = 11, colour = "#486581"),
    axis.title.x = element_math_typst(size = 13, face = "bold"),
    axis.title.y = element_math_mitex(size = 13, face = "bold"),
    legend.title = element_typst(size = 11, face = "bold"),
    legend.text = element_typst(size = 10)
  )
```

Use this family when:

- the content is part of the plot theme, not a data layer
- you want consistent Typst rendering in titles and labels
- you want Typst math and MiTeX math inside theme text

## A few practical notes

- `size.unit = "pt"` is the default in `ggtypst`; use `"mm"` when you want to
  match `ggplot2::geom_text()` sizing conventions
- `colour` and `color` are both accepted
- `face` and `fontface` are aliases
- math helpers only support `face = "plain"` and `face = "bold"`
- `inline = TRUE` is useful when you want inline rather than display-style math

## Next steps

After this article, the most useful places to continue are:

- the function reference for `annotate_*()`, `geom_*()`, and `element_*()`
- the package homepage for showcase plots
- the contributing guide if you want to understand the R/Rust architecture
